### 2024-07-02 latex/HP2_comm.tex
new input files

%%%%%%%%%%
\subsection*{2.10}
\begin{translation}[hp02_010]
If [the yogi] breathes in through Iḍā, he should then exhale the restrained [breath] through the other [channel, i.e. Piṅgalā]. Next, he should inhale through Piṅgalā, hold the breath and release it through the left [channel]. The channels of ascetics meditating on the two orbs of the sun and moon using this method are purified after three months.
\end{translation}

@@

The reading we have adopted in the third verse quarter (\emph{mukhān nirgamayec caiṣā}) is well attested but may not be original because the first hemistich has a finite verb and the \emph{eṣā} must be construed with \emph{netiḥ} in the fourth verse quarter. The alternative reading \emph{mukhanirgamanād eva} (\etaOne\ and \etaTwo) may be original but is difficult to construe and \emph{mukhān nirgamayet} makes better sense in terms of describing the final part of the practice. 
\end{philcomm}
%%%%%%%%%%
\subsection*{2.31}
\begin{translation}[hp02_031]
[The technique] with a small thread purifies the skull, bestows divine sight and quickly cures a multitude of diseases that arise above the jaw.
\end{translation}
%?? expand note on possibility of niti/nīti being original (found in alpha, betaone, epsilonone, delta i.e. in good manuscripts in most groups; vernacular?) Only niti works well in final pāda. Brahmānanda rewrites to have netī as subject, which would be best. sūtrikā is old — eta1 = V1. In 22 alphatwo has nitiś caiva...
% adopt hanūrdhva
% adopt jayaty āśu sa-sūtrikā

@@

%??% JM to supply Babaji video to be uploaded to our host. Send to Mitsuyo.

@@

%%%%%%%%%%
\subsection*{2.41}
\begin{translation}[hp02_041]
When the network of channels has been purified by breath-controls as prescribed, the breath pierces the mouth of \emph{suṣumṇā} and enters it with ease.
\end{translation}

@@

%\begin{philcomm}[hp02_045]
%\end{philcomm}

@@

The meaning of \emph{nirodhāvadhi} is not entirely clear, but all sources and the \emph{Jyotsnā} (2.49) agree on this reading. The problem is that to practise \emph{kumbhaka} ``up to cessation \emph{nirodha}'' seems to suggest that cessation is not that of the physical breath, which by definition ceases in \emph{kumbhaka}, but of the vital wind (\emph{prāṇa}) within the body. By citing a verse from an unnamed text, Brahmānanda seems to understand this verse as saying that the breath should very carefully (\emph{atiprayatnena}) be held as far as the extremities of the body so that it does not damage the body by exiting through the hair follicles:

@@

\begin{versinnote}
\tl{haṭhān niruddhaḥ prāṇo 'yaṃ romakūpeṣu niḥsaret |\\+}
\tl{dehaṃ vidārayaty eṣa kuṣṭhādi janayaty api ||\\!}%
\end{versinnote}
%??JM: how about emending to virodhāvadhi, “until it is uncomfortable”, which would make much better sense. Yes, to add to note but not emend.
\end{philcomm}

@@

\begin{versinnote}
\tl{bhavet svacchandadehas tu sarvopadravavarjitaḥ |\\+}
\tl{anena vidhinā satyaṃ yogīndro bhāti bhūtale ||\\+}
\tl{°dehas tu ] °dehaḥ syāt P; °dehasyāt T,t1\\!}
\end{versinnote}

@@

%%%%%%%%%%
\subsection*{2.67}
\begin{translation}[hp02_067]
it is an auspicious thunderbolt that awakens \emph{kuṇḍalinī}, destroys bad deeds, bestows happiness, and destroys the blockage of phlegm, etc., situated at the mouth of the central channel,[\dots]
\end{translation}

@@

Forcibly loud inhalation with the sound of a male bee; very slow exhalation with the sound of a female bee: as a result of practising thus, there arises in the minds of the best yogis an extraordinary blissful playfulness.
\end{translation}

@@

\begin{versinnote}
\tl{atha bhrāmarī---\\+}
\tl{vegodghoṣaṃ pūrakaṃ bhṛṅganādaṃ \\+}
\tl{bhṛṅgīnādaṃ recakaṃ mandamandam |\\+}
\tl{yogīndrāṇāṃ nityam abhyāsayogāc \\+}
\tl{citte jātā kā cid ānandalīlā || \\+}
\tl{līlā ]°mūrcchā N,n1,n2,n3,n4,J,T,t1\\!}
\end{versinnote}

@@

Many manuscripts of the \emph{Haṭhapradīpikā}, including the \textalpha\ group, have \emph{ānandamūrcchā} at the end of the verse, instead of \emph{ānandalīlā}, which we have adopted and which is supported by \betaTwo\ and the \textgamma\ group, as well as the most important testimonia (i.e.~the \emph{Haṭharatnāvalī}, \emph{Yogacintāmaṇi} and \emph{Yuktabhavadeva}). The reading \emph{ānanda\-mūrcchā} is probably the result of dittography as the heading \emph{atha mūrcchā} follows this verse. It has persisted in the transmission because  \emph{ānandamūrcchā} makes some sense by itself (`swooning through bliss'), although when it is read with the rest of the verse quarter, the meaning is somewhat odd (`swooning through bliss arises in the mind').
\end{philcomm}

@@

%%%%%%%%%%
\subsection*{2.70}
\begin{translation}[hp02_070]
Now mūrcchā:

At the end of inhalation [the yogi] should tightly apply the Jālandhara [lock] and exhale slowly. This, which is called \emph{mūrcchā}, bestows the bliss of  fainting.
\end{translation}
% adopt feminine in 2nd hemistich

@@

% MD: Perhaps \emph{recakaḥ pūrakaḥ kāryaḥ} which is closer to the readings of N3 and V3?  Cf. 2.74a \emph{recakaṃ pūrakaṃ muktvā}.
% JB: We need the relative pronoun to pick up on the sa in the second pāda
% MD: Not always. Cf. 1.58, 2.24, 2.26.
\end{philcomm}

@@

%%%%%%%%%%
\subsection*{2.78}
\begin{translation}[hp02_078]
At the end of exhaling the retained breath, [the yogi] should make the mind supportless. By practising in this way he reaches the state of Rājayoga .
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.79}
\begin{translation}[hp02_079]
As a result of retaining the breath, kuṇḍalinī awakens; as a result of the awakening of kuṇḍalinī, suṣumṇā becomes free of blockages and success in Haṭha arises.
\end{translation}

@@

\begin{versinnote}
\tl{kumbhakāt kuṇḍalībodhaḥ kuṇḍalībodhato bhavet |\\+}
\tl{anargalaḥ suṣumṇānto haṭhasiddhiś ca jāyate ||\\+}
\tl{anargalaḥ ] L, antargataḥ N\\!}
\end{versinnote}

@@

%%%%%%%%%%
\subsection*{2.80}
\begin{translation}[hp02_080]
Thinness of the body, clear complexion, clarity of the inner sound, very bright eyes, freedom from disease, mastery of semen, stimulation of the [body’s] fire [and] purification of the channels are the signs of success in Haṭha.
\end{translation}

>>>>

%%%%%%%%%%
\subsection*{2.10}
\begin{translation}[hp02_010]
If [the yogi] breathes in through Iḍā, he should then exhale the restrained [breath] through the other [channel, i.e. Piṅgalā]. He should inhale through Piṅgalā, hold the breath sufficiently long and release it through the left [channel]. The channels of ascetics meditating on the two orbs of the sun and moon using this method are purified after three months.
\end{translation}

%

@@

The reading we have adopted in the third verse quarter (\emph{mukhān nirgamayec caiṣā}) is well attested but may not be original because the first hemistich has a finite verb and the \emph{eṣā} must be construed with \emph{netiḥ} in the fourth verse quarter. The alternative reading \emph{mukhanirgamanād eva} (\etaOne\ and \etaTwo) may be original but is difficult to construe and \emph{mukhān nirgamayet} makes better sense in terms of describing the final part of the practice. 
\end{philcomm}
%%%%%%%%%%
\subsection*{2.31}
\begin{translation}[hp02_031]
[The technique] with a small thread purifies the skull, bestows divine sight and quickly cures a multitude of diseases that arise above the jaw.
\end{translation}
%?? Change tr, JM to write note expand note on possibility of niti/nīti being original (found in alpha, betaone, epsilonone, delta i.e. in good manuscripts in most groups; vernacular?) Only niti works well in final pāda. Brahmānanda rewrites to have netī as subject, which would be best. sūtrikā is old — eta1 = V1. In 22 alphatwo has nitiś caiva... Tricky; it seems that the spelling varies here between nīti and niti (in the header it's unanimously nīti in alpha)
% adopt hanūrdhva
% adopt jayaty āśu sa-sūtrikā

@@

\medialink{Mahant Bālyogī Rām Bālak Dās performing nauli at Dīrgheśvar Mahādev, U.P., in 2012.}{BabajiNauli.m4v}

@@

%%%%%%%%%%
\subsection*{2.41}
\begin{translation}[hp02_041]
When the network of channels has been purified by breath-controls as prescribed, the breath pierces the mouth of Suṣumṇā and enters it with ease.
\end{translation}

@@

\begin{philcomm}[hp02_045]
The term \emph{uḍḍīyāṇa} is spelt various ways among manuscripts of the \emph{Haṭhapradīpikā} and other texts. Generally speaking, the \textalpha\ group have \emph{uḍḍīyāṇa}, \textbeta\ \emph{uḍḍiyāṇa}, \textgamma\ and \textdelta\ \emph{uḍḍiyāna} and the \emph{Jyotsnā} \emph{uḍḍīyāna}. Other spellings, such as \emph{uḍyāṇa} and \emph{uḍiyāṇa}, also occur. We have adopted the spelling of \textalpha\ except in cases where the metre of a verse requires otherwise, as in 2.45d (\emph{uḍḍiyāṇakaḥ}). As noted by Sanderson (2007: 265), diverse spellings of the place by the same name occur, including Uḍḍiyāna, Oḍḍiyāna, U/Oḍiyāna, U/Oḍyāna, or U/Oḍḍayana. 
\end{philcomm}

@@

The meaning of \emph{nirodhāvadhi} is not entirely clear, but all sources and the \emph{Jyotsnā} (2.49) agree on this reading. The original reading may well have been \emph{virodhāvadhi}, “until it is uncomfortable”, which makes better sense but we have decide not to emend accordingly. The problem with \emph{nirodhāvadhi} is that to practise \emph{kumbhaka} ``up to cessation \emph{nirodha}'' seems to suggest that cessation is not that of the physical breath, which by definition ceases in \emph{kumbhaka}, but of the vital wind (\emph{prāṇa}) within the body. By citing a verse from an unnamed text, Brahmānanda seems to understand this verse as saying that the breath should very carefully (\emph{atiprayatnena}) be held as far as the extremities of the body so that it does not damage the body by exiting through the hair follicles:

@@

\begin{versinnote}
\tl{haṭhān niruddhaḥ prāṇo 'yaṃ romakūpeṣu niḥsaret |\\+}
\tl{dehaṃ vidārayaty eṣa kuṣṭhādi janayaty api ||\\!}%
\end{versinnote}
\end{philcomm}

@@

\begin{versinnote}
\tl{bhavet svacchandadehas tu sarvopadravavarjitaḥ |\\+}
\tl{anena vidhinā satyaṃ yogīndro bhāti bhūtale ||\\+}
\tl{\var{°dehas tu ] °dehaḥ syāt P; °dehasyāt T,t1}\\!}
\end{versinnote}

@@

%%%%%%%%%%
\subsection*{2.67}
\begin{translation}[hp02_067]
it is an auspicious thunderbolt that awakens Kuṇḍalinī, destroys bad deeds, bestows happiness, and destroys the blockage of phlegm, etc., situated at the mouth of the central channel,[\dots]
\end{translation}

@@

Forcibly loud inhalation with the sound of a male bee; very slow exhalation with the sound of a female bee: as a result of practising thus, there arises in the minds of the best yogis a certain blissful stupefaction.
\end{translation}

@@

\begin{versinnote}
\tl{atha bhrāmarī---\\+}
\tl{vegodghoṣaṃ pūrakaṃ bhṛṅganādaṃ \\+}
\tl{bhṛṅgīnādaṃ recakaṃ mandamandam |\\+}
\tl{yogīndrāṇāṃ nityam abhyāsayogāc \\+}
\tl{citte jātā kā cid ānandalīlā || \\+}
\tl{\var{līlā ]°mūrcchā N,n1,n2,n3,n4,J,T,t1}\\!}
\end{versinnote}

@@

Witnesses of the \gamma \ and \delta \ branches of the stemma, and Brahmānanda, have \emph{ānandalīlā} instead of \emph{ānandamūrchā}, perhaps in order to avoid the repetition of \emph{mūrchā}, which is also found in the following verse.

%Old note before mūrcchā adopted. Many manuscripts of the \emph{Haṭhapradīpikā}, including the \textalpha\ group, have \emph{ānandamūrchā} at the end of the verse, instead of \emph{ānandalīlā}, which we have adopted and which is supported by \betaTwo\ and the \textgamma\ group, as well as the most important testimonia (i.e.~the \emph{Haṭharatnāvalī}, \emph{Yogacintāmaṇi} and \emph{Yuktabhavadeva}). The reading \emph{ānanda\-mūrcchā} is probably the result of dittography as the heading \emph{atha mūrchā} follows this verse. It has persisted in the transmission because  \emph{ānandamūrchā} makes some sense by itself (`swooning through bliss'), although when it is read with the rest of the verse quarter, the meaning is somewhat odd (`swooning through bliss arises in the mind').
\end{philcomm}

@@

%%%%%%%%%%
\subsection*{2.70}
\begin{translation}[hp02_070]
Now mūrchā:

At the end of inhalation [the yogi] should tightly apply the Jālandhara [lock] and exhale slowly. This fainting of the mind, which is called \emph{mūrchanā}, gives pleasure.
\end{translation}
%name is given as mūrchanā but headings in alpha etc. are mūrchā. Some, including epsilon1 do have mūrchanā.
% adopt feminine in 2nd hemistich
% JB: the meaning of manomūrcchā is not conveyed by the current translation. Fainting of the mind is different to normal fainting where a person falls to the ground unconsciousness. It's unlikely that the person in mūrcchā kumbhaka was supposed to collapse.... bestows bliss, when the mind faints.

@@

% MD: Perhaps \emph{recakaḥ pūrakaḥ kāryaḥ} which is closer to the readings of N3 and V3?  Cf. also 2.74a \emph{recakaṃ pūrakaṃ muktvā}.
% JB: We need the relative pronoun to pick up on the sa in the second pāda
% MD: Not necessarily. Cf. 1.58, 2.24, 2.26.
\end{philcomm}

@@

%%%%%%%%%%
\subsection*{2.78}
\begin{translation}[hp02_078]
At the end of exhaling the retained breath, [the yogi] should make the mind supportless. By practising in this way he reaches the state of Rājayoga.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.79}
\begin{translation}[hp02_079]
As a result of retaining the breath, Kuṇḍalinī awakens; as a result of the awakening of Kuṇḍalinī, Suṣumṇā becomes free of blockages and success in Haṭha arises.
\end{translation}

@@

\begin{versinnote}
\tl{kumbhakāt kuṇḍalībodhaḥ kuṇḍalībodhato bhavet |\\+}
\tl{anargalaḥ suṣumṇānto haṭhasiddhiś ca jāyate ||\\+}
\tl{\var{anargalaḥ ] L, antargataḥ N}\\!}
\end{versinnote}

@@

%%%%%%%%%%
\subsection*{2.80}
\begin{translation}[hp02_080]
Thinness of the body, clear complexion, clarity of the inner sound,  bright eyes, freedom from disease, mastery of semen, stimulation of the [body’s] fire and purification of the channels are the signs of success in Haṭha.
\end{translation}
%MD: Pāda d: from a stemmatic point of view, nāḍīṣu śuddhir is better than nāḍīviśuddhir.JM: adopt
% and Pāda b: ca nirmale is better than sunirmale. JM: adopt
